Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox

### 2021-08-10 manuscript/elife/elife-manuscript.tex
more reviewer feedback incorporated

Finally, the poor fit between observed and predicted levels of diversity across
species is not remedied by stronger selection parameters. In
\FIGSUPP[linksel]{linked-sel-alt}B, I increase both selection parameters $U$
and $\gamma$ ten-fold each, and find the same qualitative pattern: on a log-log
scale the relationship between $N_c$ and $\pi$ is linear, while the predicted
diversity under linked selection is non-linear with $N_c$. Under this ten-fold
higher selection regime, there is more overlap between observed and predicted
levels of diversity, but diversity is severely \emph{under-predicted} for
high-$N_c$ species. Additionally, this would imply that selection in
low-to-mid-$N_c$ species is ten-folder higher than estimated in
\emph{Drosophila melanogaster}, which is implausible.

Overall, predicted levels of diversity under models of recurrent hitchhiking
and background selection fit the observed relationship between $N_c$ and $\pi$
poorly, and this is not overly sensitive to parameter choice. These findings
extend the argument of \cite{Coop2016-gx}, showing that not only are previous
empirical estimates of the impact of linked selection unable to explain
Lewontin's Paradox, but it is unlikely current linked selection models under a
wide range of selection parameters are capable of explaining the observed
patterns of diversity across species.

@@

In this study, my goal was not to accurately estimate the levels in diversity
across species, but rather to give linked selection the best possible chance to
solve Lewontin's Paradox. Still, I find that even after parameterizing
hitchhiking and background selection with strong selection parameter estimates
from \emph{Drosophila melanogaster}, the predicted patterns of diversity under
linked selection poorly fit observed patterns of diversity across species.
This result extends the analysis by \cite{Coop2016-gx} showing that levels
of $\pi_0$ estimated by \cite{Corbett-Detig2015-gt} are not decoupled from
genome-wide average $\pi$, as would occur if linked selection were to explain
Lewontin's Paradox. Here, my analysis goes a step further and suggests that
models of recurrent hitchhiking and background selection are not capable of
explaining the observed relationship between $\pi$ and census size, in part
because mid-$N_c$ species have sufficiently long recombination map lengths to
diminish the effects of even strong selection. This finding supports the idea
the levels of diversity across species are primarily determined by past
demographic fluctuations. Overall, while this suggests these two common modes
of linked selection seem unlikely to explain across-taxa patterns of diversity,
there are three major potential limitations of my approach that need further
evaluation.

First, I approximate the reduction in diversity using homogeneous background
selection and recurrent hitchhiking models
\citep{Kaplan1989-sc,Hudson1995-xc,Coop2012-cd}, when in reality, there is
genome-wide heterogeneity in functional density, recombination rates, and the
adaptive substitutions across species. Each of these factors mediate how
strongly linked selection impacts diversity across the genome. Despite these
model simplifications, the predicted reduction in diversity in \emph{Drosophila
melanogaster} is $85\%$ (when using $N_e$, not $N_c$), which is reasonably
close to the estimated $77\%$ from the more realistic model of Elyashiv et al.
that accounts for the actual position of substitutions, annotation features,
and recombination rate heterogeneity (though it should be noted that these both
use the same parameter estimates). Furthermore, even though my model fails to
capture the heterogeneity of functionality density and recombination rate in
real genomes, it is still extraordinary conservative, likely overestimating the
effects of linked selection to see if it could be capable of decoupling
diversity from census size and explain Lewontin's Paradox. This is in part
because the strong selection parameter estimates from \emph{Drosophila
melanogaster} used, but also because I assume that the effective population
size is equal to the census size.  Even then, this decoupling only occurs in
very high--census-size species, and implies that the diversity in the absence
of linked selection, $\pi_0$, is currently underestimated by several orders of
magnitude. Moreover, the study of \cite{Corbett-Detig2015-gt} did consider
recombination rate and functional density heterogeneity in estimating the
reduction due to linked selection across species, yet their predicted
reductions are orders of magnitude weaker than those considered here by
assuming that $N_e = N_c$ (\FIGSUPP[linksel]{reductions}B).
Overall, even with more realistic models of linked selection, current models of
linked selection seem fundamentally unable to fit the diversity--census-size
relationship.

Second, my model here only considers hard sweeps, and ignores the contribution
of soft sweeps (e.g. from standing variation or recurrent mutations;
\cite{Hermisson2005-hs,Pennings2006-lj}), partial sweeps (e.g those that do not
reach fixation), and the interaction of sweeps and spatial processes. While
future work exploring these alternative types of sweeps is needed, the
predicted reductions in diversity found here under the simplified sweep model
are likely relatively robust to these other modes of sweeps for a few reasons.
First, the shape of the diversity--recombination curve is equivalent under
models of partial sweeps and hard sweeps, though these imply different rates of
sweeps \citep{Coop2012-cd}. Second, in the limit where most fitness
variation is due to weak soft sweeps from standing variation scattered across
the genome (i.e. due to polygenic fitness variation), levels of diversity are
well approximated by quantitative genetic linked selection models
\citep{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}. The reduction in
diversity under these models is nearly identical to that under background
selection models, in part because deleterious alleles at mutation-selection
balance constitute a considerable component of fitness variation (see Appendix
Section \ref{app:bgs}; \cite{Charlesworth2000-km,Charlesworth2015-am}). Third,
the parameters from \cite{Elyashiv2016-vt} are robust to many types of
sweeps that result in substitution (e.g. see p. 19 of their Supplementary
Online Materials). Finally, I also disregarded the interaction of sweeps and
spatial processes.  For populations spread over wide ranges, limited dispersal
slows the spread of sweeps, allowing for new beneficial alleles to arise,
spread, and compete against other segregating beneficial variants
\citep{Ralph2015-kl,Ralph2010-ki}. Through limited dispersal should act to
``soften sweeps" and not impact my findings for the reasons described above,
future work could investigate how these processes impact diversity in ways not
captured by hard sweep models.

Third, other selective processes, such as fluctuating selection or hard
selective events, could reduce diversity in ways not captured by the background
selection and hitchhiking model. Since frequency-independent fluctuating
selection generally reduces diversity under most conditions
\citep{Novak2017-yp}, this could lead seasonality and other sources of
temporal heterogeneity to reduce diversity in large-$N_c$ species with short
generation times more than longer-lived species with smaller population sizes.
Future work could consider the impact of fluctuating selection on diversity
under simple models \citep{Barton2000-zg} if estimates of key parameters
governing the rate of such fluctuations were known across taxa. Additionally,
another mode of selection that could severely reduce diversity across taxa, yet
remains unaccounted for in this study, is periodic hard selective events. These
selective events could occur regularly in a species' history yet be
indistinguishable from demographic bottlenecks with just population genomic
data.

@@

One limitation of this study is the inability to quantify the impact of spatial
population genetic processes on the relationship between diversity and census
population sizes across taxa. The genomic diversity estimates collated in this
study unfortunately lack details about the sampling process and spatial data,
which can have a profound impact on population genomic summary statistics
\citep{Battey2020-lc}. These issues could systematically bias species-wide
diversity estimates; for example, if diversity estimates from a cosmopolitan
species were primarily from a single subpopulation, diversity would be an
underestimate relative to the entire population. However, biased spatial
sampling alone seems incapable of explaining the $\pi$-$N_c$ divergence in
high-$N_c$ taxa. In the extreme scenario in which only one subpopulation was
sampled, $F_\text{ST}$ would need to be close to one for population subdivision
alone to sufficiently reduce the total population heterozygosity to explain the
orders-of-magnitude shortfall between predicted and observed diversity levels.
This is because the equation for $F_{ST}$ can be rearranged such that $H_S =
(1-F_{ST})H_T$, where $H_S$ and $H_S$ are the subpopulation and total
population heterozygosities; if $H_T = 4N_c\mu$, then only $F_{ST} \approx 1$
can reduce $H_S$ several orders of magnitude. Yet, across-taxa surveys indicate
that $F_\text{ST}$ is almost never this high within species
\citep{Roux2016-lm}.  Still, future work could quantify the extent to which
spatial processes contribute to Lewontin's Paradox. For example, high-$N_c$
taxa usually experience range expansions, likely with repeated founder effects
and local extinction/recolonization dynamics that doubtlessly depress
diversity. In particular, with the appropriate data, one could estimate the
empirical relationship between dispersal distance, range size, and coalescent
effective population size across taxa.

In this study, I have focused entirely on assessing the role of linked
selection, rather than demography, in reducing diversity across taxa. In
contrast to demographic models, models of linked selection have comparatively
fewer parameters and more readily permit rough estimates of diversity
reductions across taxa. Still, a full resolution of Lewontin's Paradox would
require understanding how the demographic processes across taxa with incredibly
heterogeneous ecologies and life histories transform $N_c$ into $N_e$. With
population genomic data becoming available for more species, this could involve
systematically inferring the demographic histories of tens of species and
looking for correlations in the frequency and size of bottlenecks with $N_c$
across species.

>>>>

Finally, the poor fit between observed and predicted levels of diversity across
species is not remedied by stronger selection parameters. In
\FIGSUPP[linksel]{linked-sel-alt}B, I increase both selection parameters $U$
and $\gamma$ ten-fold each, and find the same qualitative pattern: on a log-log
scale the relationship between $N_c$ and $\pi$ is linear, while the predicted
diversity under linked selection is non-linear with $N_c$. Under this ten-fold
higher selection regime, there is more overlap between observed and predicted
levels of diversity, but diversity is severely \emph{under-predicted} for
high-$N_c$ species. Additionally, this would imply that selection in
low-to-mid-$N_c$ species is ten-folder higher than estimated in
\emph{Drosophila melanogaster}, which is implausible. Overall, predicted
levels of diversity under models of recurrent hitchhiking and background
selection fit the observed relationship between $N_c$ and $\pi$ poorly, and
this is not overly sensitive to parameter choice.

@@

Even assuming selection parameters estimated from \emph{Drosophila
melanogaster}, where the effects of linked selection are thought to be
especially strong, the predicted patterns of diversity under linked selection
poorly fit observed patterns of diversity across species. My results support
the analysis by \cite{Coop2016-gx} showing that levels of $\pi_0$ estimated by
\cite{Corbett-Detig2015-gt} are not decoupled from genome-wide average $\pi$,
as would occur if linked selection were to explain Lewontin's Paradox.
Additionally, my analysis goes a step further, showing that current linked
selection models under a wide range of selection parameters are incapable of
explaining the observed relationship between census size and diversity. This is
in part because mid-$N_c$ species have sufficiently long recombination map
lengths to diminish the effects of even strong selection. Overall, while this
suggests these two common modes of linked selection seem unlikely to explain
patterns of diversity across taxa, there are three major potential limitations
of my approach that need further evaluation.

First, I approximate the reduction in diversity using homogeneous background
selection and recurrent hitchhiking models
\citep{Kaplan1989-sc,Hudson1995-xc,Coop2012-cd}, when in reality, there is
genome-wide heterogeneity in functional density, recombination rates, and the
adaptive substitutions across species. Each of these factors mediate how
strongly linked selection impacts diversity across the genome. Despite these
model simplifications, the predicted reduction in diversity in \emph{Drosophila
melanogaster} is $85\%$ (when using $N_e$, not $N_c$), which is reasonably
close to the estimated $77\%$ from the more realistic model of Elyashiv et al.
that accounts for the actual position of substitutions, annotation features,
and recombination rate heterogeneity (though it should be noted that these both
use the same parameter estimates). Furthermore, even though my model fails to
capture the heterogeneity of functionality density and recombination rate in
real genomes, it is still conservative, likely overestimating the effects of
linked selection to see if it could be capable of decoupling diversity from
census size and explain Lewontin's Paradox. This is in part because the strong
selection parameter estimates from \emph{Drosophila melanogaster} used, but
also because I assume that the effective population size is equal to the census
size. Even then, this decoupling only occurs in very high--census-size species,
and implies that the diversity in the absence of linked selection, $\pi_0$, is
currently underestimated by several orders of magnitude. Moreover, the study of
\cite{Corbett-Detig2015-gt} did consider recombination rate and functional
density heterogeneity in estimating the reduction due to linked selection
across species, yet their predicted reductions are orders of magnitude weaker
than those considered here by assuming that $N_e = N_c$
(\FIGSUPP[linksel]{reductions}B). Overall, given the effects estimated under
more realistic inference models are still orders of magnitude weaker than those
used in this study, current models of linked selection seem fundamentally
unable to fit the diversity--census-size relationship.

Second, my model here only considers hard sweeps, and ignores the contribution
of soft sweeps (e.g. from standing variation or recurrent mutations;
\cite{Hermisson2005-hs,Pennings2006-lj}), partial sweeps (e.g those that do not
reach fixation), and the interaction of sweeps and spatial processes. While
future work exploring these alternative types of sweeps is needed, the
predicted reductions in diversity found here under the simplified sweep model
are likely relatively robust to these other modes of sweeps for a few reasons.
First, the shape of the diversity--recombination curve is equivalent under
models of partial sweeps and hard sweeps, though these imply different rates of
sweeps \citep{Coop2012-cd}. Second, in the limit where most fitness variation
is due to weak soft sweeps from standing variation scattered across the genome
(i.e. due to polygenic fitness variation), levels of diversity are well
approximated by quantitative genetic linked selection models
\citep{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}. The reduction in
diversity under these models is nearly identical to that under background
selection models, in part because deleterious alleles at mutation-selection
balance constitute a considerable component of fitness variation (see Appendix
Section \ref{app:bgs}; \cite{Charlesworth2000-km,Charlesworth2015-am}). Third,
the parameters from \cite{Elyashiv2016-vt} could reflect a mixture of types of
sweeps (\cite{Elyashiv2016-vt} p. 14 and p. 19 of their Supplementary Online
Materials). Finally, I also disregarded the interaction of sweeps and spatial
processes.  For populations spread over wide ranges, limited dispersal slows
the spread of sweeps, allowing for new beneficial alleles to arise, spread, and
compete against other segregating beneficial variants
\citep{Ralph2015-kl,Ralph2010-ki}. Through limited dispersal should act to
``soften sweeps" and not impact my findings for the reasons described above,
future work could investigate how these processes impact diversity in ways not
captured by hard sweep models.

Third, other selective processes, such as fluctuating selection or hard
selective events (i.e. selection resulting in a reduction in the population
size), could reduce diversity in ways not captured by the background selection
and hitchhiking models. Since frequency-independent fluctuating selection
reduces diversity under most conditions \citep{Novak2017-yp}, this could lead
seasonality and other sources of temporal heterogeneity to reduce diversity in
large-$N_c$ species with short generation times more than longer-lived species
with smaller population sizes.  Future work could consider the impact of
fluctuating selection on diversity under simple models \citep{Barton2000-zg} if
estimates of key parameters governing the rate of such fluctuations were known
across taxa. Additionally, another mode of selection that could severely reduce
diversity across taxa, yet remains unaccounted for in this study, is periodic
hard selective events. These selective events could occur regularly in a
species' history yet be indistinguishable from demographic bottlenecks with
just population genomic data.

@@

One limitation of this study is the inability to quantify the impact of spatial
and demographic population genetic processes on the relationship between
diversity and census population sizes across taxa. The genomic diversity
estimates collated in this study unfortunately lack details about the sampling
process and spatial data, which can have a profound impact on population
genomic summary statistics \citep{Battey2020-lc}. These issues could
systematically bias species-wide diversity estimates; for example, if diversity
estimates from a cosmopolitan species were primarily from a single region or
subpopulation, diversity would be an underestimate relative to the entire
population. However, biased spatial sampling alone seems incapable of
explaining the $\pi$-$N_c$ divergence in high-$N_c$ taxa. In the extreme
scenario in which only one subpopulation was sampled, $F_\text{ST}$ would need
to be close to one for population subdivision alone to sufficiently reduce the
total population heterozygosity to explain the orders-of-magnitude shortfall
between predicted and observed diversity levels.  This can be seen by
rearranging the expression for $F_{ST}$ as $H_S = (1-F_{ST})H_T$, where $H_S$
and $H_T$ are the subpopulation and total population heterozygosities; if $H_T
= 4N_c\mu$, then only $F_{ST} \approx 1$ can reduce $H_S$ several orders of
magnitude. Yet, across-taxa surveys indicate that $F_\text{ST}$ is almost never
this high within species \citep{Roux2016-lm}. Future work could quantify the
extent to which more realistic spatial processes contribute to Lewontin's
Paradox. For example, high-$N_c$ taxa usually experience range expansions, with
repeated founder effects and local extinction/recolonization dynamics that
depress diversity \citep{Slatkin1977-kd}. In particular, with the appropriate
data, one could estimate the empirical relationship between dispersal distance,
range size, and coalescent effective population size across taxa.

In this study, I have focused entirely on assessing the role of linked
selection, rather than demography, in reducing diversity across taxa. In
contrast to demographic models, models of linked selection have comparatively
fewer parameters and more readily permit rough estimates of diversity
reductions across taxa.  Given that I find that models of linked selection are
incapable of explaining the observed relationship between $N_c$ and $\pi$, this
supports the hypothesis the diversity across species are shaped primarily by
past demographic fluctuations. Still, a full resolution of Lewontin's Paradox
would require understanding how the demographic processes across taxa with
incredibly heterogeneous ecologies and life histories transform $N_c$ into
$N_e$. With population genomic data becoming available for more species, this
could involve systematically inferring the demographic histories of tens of
species and looking for correlations in the frequency and size of bottlenecks
with $N_c$ across species.
